Exploring dimensions of biodiversity in Japanese ferns

### 2021-08-20 ms/manuscript.Rmd
Update MS
according to comments from WI
Update reference processing
Combine other references and main_library before filtering
Count plastome sampling

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "repro_data",
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  "plastome_calibration_dates",
  "plastome_tree",
  "green_list",
  "model_params",
  "lrt_results",
  "aic_env_repro",
  "model_fits",
  "biodiv_ferns_cent",
  "refs_yaml",
  "refs_other_yaml",
  "template_file",
  "csl_file"
))
```

@@

@@

We downloaded climate data at the 2.5 minute scale from the WorldClim database using the `r func("getData")` function in the R package `r pack("raster")`, and extracted four variables related to our research questions as follows: mean annual temperature (bio1; hereafter "temperature"), temperature seasonality (bio4), annual precipitation (bio12; hereafter "precipitation"), and precipitation seasonality (bio15) (`r s_figure("envir")`). 
We identified the intersection of the climate data at the 2.5 minute scale with the 0.2° grid-cells, then calculated the mean of each climatic variable for each 0.2° grid-cell.
We calculated area as rolling mean area (km^2^) across 1 degree latitudinal bands.
This approach reflects the idea that the amount of area with a given climate regime enclosing each grid cell may influence species richness of that cell [@Fahrig2013], and that using raw island areas would not be appropriate in this context as there are just four very large islands with many much smaller ones located close by.
The R package `r pack("sf")` was used for GIS analyses [@Pebesma2018].

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]. 
However, using this dataset alone for phylogeny suffers two drawbacks: it is not densely sampled enough to resolve many internal nodes, and it lacks many lineages that have fossils available for molecular dating. 
Because one goal of the study is to understand the distribution of paleo- vs. neo-endemic areas (defined in units of time), we require an ultrametric tree. 
To obtain a maximally sampled, robust phylogenetic tree, we combined the *rbcL* data of @Ebihara2019b with a globally sampled plastid dataset including all fern sequences on GenBank for four plastid genes (*rbcL*, *atpB*, *aptA*, and *rps4*) and 66 other coding, single-copy plastid genes extracted from 100 complete fern plastomes (Nitta et al., in prep) (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` OTUs total). 
We conducted phylogenetic analysis with `r software("IQTREE")` [@Nguyen2015] and molecular dating with `r software("treePL")` [@Smith2012] using `r nrow(plastome_calibration_dates)` fossil calibration points after @Testo2016a. 
We then trimmed the tree to Japanese taxa only and used this for all further analyses.
Tree visualization was done with the `r pack("ggtree")` R package [@Yu2017].

@@

We measured functional diversity (FD) following the method of @Petchey2002. 
We first calculated functional distances using Gower's method [@Gower1971] on the trait data with the `r func("gowdis")` function in the R package `r pack("FD")` [@Laliberte2010]. 
We weighted continuous and qualitative traits equally, weighted each binary component trait of each qualitative trait equally, and log-transformed and scaled continuous traits prior to calculating distances. 
We then used the distances to build a trait dendrogram using the `r func("hclust")` function in the `r pack("stats")` R package [@RCoreTeam2020], and used the dendogram in place of a phylogenetic tree to calculate PD and RPD as described above. 
These functional analogs of PD and RPD are hereafter referred to as FD (functional diversity) and RFD (relative functional diversity).

@@

We first tested for spatial autocorrelation in biodiversity metrics (richness, SES of PD, SES of FD, SES of RPD, SES of RFD) and independent variables (environmental variables and % apomictic taxa) using Moran's I with the `r func("moran.mc")` function in the `r pack("spdep")` package. 
In all cases, significant spatial autocorrelation was detected (Moran's I > 0, *P* value of permutation test < 0.05) (`r s_table("moran")`). 
Therefore, we used spatially explicit methods for modeling.

We checked for correlation between independent variables while taking into account spatial autocorrelation using a modified *t*-test with the `r func("modified.ttest")` function in the `r pack("SpatialPack")` R package. 
This indicated that temperature seasonality and % apomictic taxa were correlated with mean annual temperature (`r s_table("corr")`), so for the main modeling analysis we only included the following environmental variables: mean annual temperature, precipitation, and precipitation seasonality.

@@

Author contributions: J.H.N., B.M., and A.E. conceived the ideas; A.E. provided the data; J.H.N. analyzed the data; I.W. provided computational resources; J.H.N. wrote the manuscript with input from all co-authors.

>>>>

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "repro_data",
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  "plastome_calibration_dates",
  "plastome_tree",
  "green_list",
  "model_params",
  "lrt_results",
  "aic_env_repro",
  "model_fits",
  "biodiv_ferns_cent",
  "refs_yaml",
  "refs_other_yaml",
  "template_file",
  "csl_file",
  "broad_alignment_list",
  "japan_rbcL"
))
```

@@

All computational analyses were carried out in R v.`r getRversion()` [@RCoreTeam2020] unless otherwise stated. The `r pack("targets")` R package was used to maintain analysis workflow [@Landau2021].

@@

We downloaded climate data at the 2.5 minute scale from the WorldClim database using the `r func("getData")` function in the R package `r pack("raster")` [@R-raster], and extracted four variables related to our research questions as follows: mean annual temperature (bio1; hereafter "temperature"), temperature seasonality (bio4), annual precipitation (bio12; hereafter "precipitation"), and precipitation seasonality (bio15) (`r s_figure("envir")`). 
We identified the intersection of the climate data at the 2.5 minute scale with the 0.2° grid-cells, then calculated the mean of each climatic variable for each 0.2° grid-cell.
We calculated area as rolling mean area (km^2^) across 1 degree latitudinal bands.
This approach reflects the idea that the amount of area with a given climate regime enclosing each grid cell may influence species richness of that cell [@Fahrig2013], and that using raw island areas would not be appropriate in this context as there are just four very large islands with many much smaller ones located close by.
The R package `r pack("sf")` was used for GIS analyses [@Pebesma2018].

@@

```{r phylo-analysis-stats}
# Calculate stats related to Fern Tree of Life (FTOL) dataset

# Define outgroup taxa
plastome_og <-
  tibble(
    taxon = c(
      "Physcomitrium_patens",
      "Marchantia_polymorpha",
      "Anthoceros_angustus",
      "Isoetes_virginica",
      "Lycopodium_clavatum",
      "Selaginella_moellendorffii",
      "Sciadopitys_verticillata",
      "Ginkgo_biloba",
      "Kadsura_longipedunculata",
      "Trithuria_inconspicua",
      "Amborella_trichopoda",
      "Acorus_calamus",
      "Magnolia_tripetala",
      "Lamprocapnos_spectabilis",
      "Plumbago_auriculata",
      "Nicotiana_tabacum",
      "Liquidambar_formosana",
      "Sorbus_torminalis"),
    taxon_status = "outgroup")

# Define Sanger-sequenced genes
sanger_genes <- c("atpA", "atpB", "rbcL", "rps4")

# Make gene x taxon table for plastome alignment
plastome_sampling <-
  tibble(
    gene = names(broad_alignment_list),
    taxon = map(broad_alignment_list, rownames)
  ) %>%
  unnest(cols = taxon) %>%
  # Check that all Sanger genes are included
  verify(all(sanger_genes %in% gene)) %>%
  # Check that all outgroup taxa are included
  verify(all(plastome_og$taxon %in% taxon)) %>%
  # Categorize genes into Sanger or other
  mutate(data_type = if_else(
    gene %in% sanger_genes,
    "sanger",
    "full_plastome"
  ))

# Number of full plastome genes in FTOL dataset
n_plastome_genes <- plastome_sampling %>%
  filter(data_type == "full_plastome") %>%
  pull(gene) %>%
  n_distinct() %>%
  number(big.mark = ",")

# Number of fern taxa with full plastome sequences in FTOL dataset
n_plastome_fern_taxa <- plastome_sampling %>%
  filter(data_type == "full_plastome") %>%
  # Exclude outgroups
  anti_join(plastome_og, by = "taxon") %>%
  pull(taxon) %>%
  n_distinct() %>%
  number(big.mark = ",")

# Number of fern taxa in FTOL dataset
n_ftol_fern_taxa <- plastome_sampling %>%
  # Exclude outgroups
  anti_join(plastome_og, by = "taxon") %>%
  pull(taxon) %>%
  n_distinct() %>%
  number(big.mark = ",")

# Number of taxa in Japan fern rbcL dataset
n_japan_fern_taxa <- japan_rbcL %>% 
  rownames %>% 
  n_distinct() %>%
  number(big.mark = ",")

# Verify that tips in the plastome tree are accounted for
# by FTOL sampling + japan fern rbcL
tibble(taxon = plastome_tree$tip.label) %>%
  anti_join(select(plastome_sampling, taxon) %>% unique) %>%
  anti_join(tibble(taxon = rownames(japan_rbcL))) %>%
  verify(nrow(.) == 0, success_fun = success_logical)
```

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]. 
However, using this dataset alone for phylogeny suffers two drawbacks: it is not densely sampled enough to resolve many internal nodes, and it lacks many lineages that have fossils available for molecular dating. 
Because one goal of the study is to understand the distribution of paleo- vs. neo-endemic areas (defined in units of time), we require an ultrametric tree. 
To obtain a maximally sampled, robust phylogenetic tree, we combined the *rbcL* data of @Ebihara2019b (`r n_japan_fern_taxa` taxa) with a globally sampled plastid dataset including all fern sequences on GenBank for four widely-sequenced plastid genes (*atpA*, *atpB*, *rbcL*, and *rps4*; `r n_ftol_fern_taxa` taxa) and `r n_plastome_genes` other coding, single-copy plastid genes extracted from `r n_plastome_fern_taxa` complete fern plastomes (Nitta et al., in prep.) (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` OTUs total, including outgroup taxa). 
We conducted phylogenetic analysis with `r software("IQTREE")` [@Nguyen2015] and molecular dating with `r software("treePL")` [@Smith2012] using `r nrow(plastome_calibration_dates)` fossil calibration points after @Testo2016a. 
We then trimmed the tree to Japanese taxa only and used this for all further analyses.
Tree visualization was done with the `r pack("ggtree")` R package [@Yu2017].

@@

We measured functional diversity (FD) following the method of @Petchey2002. 
We first calculated functional distances using Gower's method [@Gower1971] on the trait data with the `r func("gowdis")` function in the R package `r pack("FD")` [@Laliberte2010]. 
We weighted continuous and qualitative traits equally, weighted each binary component trait of each qualitative trait equally, and log-transformed and scaled continuous traits prior to calculating distances.
We also tested alternative weighting schemes, but these produced very similar results (data not shown).
We then used the distances to build a trait dendrogram using the `r func("hclust")` function in the `r pack("stats")` R package [@RCoreTeam2020], and used the dendogram in place of a phylogenetic tree to calculate PD and RPD as described above. 
These functional analogs of PD and RPD are hereafter referred to as FD (functional diversity) and RFD (relative functional diversity).

@@

We first tested for spatial autocorrelation in biodiversity metrics (richness, SES of PD, SES of FD, SES of RPD, SES of RFD) and independent variables (environmental variables and % apomictic taxa) using Moran's I with the `r func("moran.mc")` function in the `r pack("spdep")` package [@Bivand2013]. 
In all cases, significant spatial autocorrelation was detected (Moran's I > 0, *P* value of permutation test < 0.05) (`r s_table("moran")`). 
Therefore, we used spatially explicit methods for modeling.

We checked for correlation between independent variables while taking into account spatial autocorrelation using a modified *t*-test with the `r func("modified.ttest")` function in the `r pack("SpatialPack")` R package [@SpatialPack2020]. 
This indicated that temperature seasonality and % apomictic taxa were correlated with mean annual temperature (`r s_table("corr")`), so for the main modeling analysis we only included the following environmental variables: mean annual temperature, precipitation, and precipitation seasonality.

@@

Author contributions: J.H.N., B.M., and A.E. conceived the ideas; A.E. provided the data; J.H.N. analyzed the data; W.I provided resources; J.H.N. wrote the manuscript with input from all co-authors.
